%% ---------------------------------------------------------------------------------------------------------------------

\chapter{Conclusion}\label{ch:conclusion}

In this thesis, a detailed study of the \unsafe{} \acrshort{API} provided by Google's Go programming language with
respect to a security context was conducted.
A thorough analysis of the dangers that come with its usage was presented, including proof-of-concept exploits that show
how possible vulnerabilities can be exploited in practice.
This analysis particularly looked at \checkNum{three} main areas of dangers.
Types with platform-dependent types such as \textit{int} can cause problems when they are converted to other types
directly, because the underlying memory might not align when the project is compiled for a different architecture, or
the memory layout of the types after compilation is changed in future versions of Go.
Incorrect conversions of slices through the internal slice representation used by the Go runtime can cause the garbage
collector to miss a live value, or the Go compiler's escape analysis algorithm to misplace a value on the stack rather
than the heap, both of which can cause non-deterministic or deterministic use-after-free vulnerabilities, respectively.
Buffer overflow vulnerabilities, which can be introduced by constructing slice values from unsuitable buffers, can lead
to severe consequences such as information leak or code injection vulnerabilities.

Thus, it is needed to audit usages of the \unsafe{} API in projects.
Such usages can be introduced through dependencies, in fact most are.
Therefore, dependencies must be checked, too.
The novel static code analysis tool \toolGeiger{}, which can help developers with this task, was presented.
It can identify and count \unsafe{} usages in Go packages, including their dependencies.
This allows developers to focus their \unsafe{} audit efforts to the most sensible packages.
Such a tool was previously available for other languages, but not for Go.
Using \toolGeiger{}, an empirical study on the current state of \unsafe{} usage in \projsAnalyzed{} of the top
\projsTotal{} most-starred open-source Go projects on \github{} was presented.
This study revealed that \percentageUnsafeProjects{} of the projects contain \unsafe{} usages, however
\percentageUnsafeTransitiveWithDependencies{} include \unsafe{} code through third-party dependencies.
Of the \packagesAnalyzed{} packages that were analyzed in total, \percentageUnsafePackages{} contained \unsafe{} code.
The average depth in the dependency tree of \averageUnsafeImportDepth{} showed that most packages with \unsafe{} usages
are reasonably close to the root module.
However, it is still hard to manually audit the complete code base including external libraries, highlighting the
importance of support by developer tools for this task.

A novel data set of \numberLabeledCodeSnippets{} manually labeled \unsafe{} code samples was created.
They are classified in two dimensions, on what is being done and for what purpose.
This data set shows that the the most common reasons for using \unsafe{} are optimizations and efficiency,
interoperability with external libraries, or to circumvent language limitations.

Furthermore, \toolSafer{} was presented, a novel \toolVet{}-style linter that is focused around \unsafe{} code.
It can identify \checkNum{two} dangerous and common usage patterns: incorrect conversions of slice and string header
values to actual slices or strings, which can introduce use-after-free vulnerabilities, and in-place conversions of
types that have platform-dependent sizes or memory layouts.
The performance of \toolSafer{} was evaluated both on the new labeled data set of \unsafe{} usages and on a set of
open-source Go packages, selected by their number of \unsafe{} usages and lines of code, which were reviewed manually.
In this evaluation, \toolSafer{} achieved excellent results, with an accuracy of
\goSaferEvaluationDatasetGosaferAccuracy{} on the labeled data set and \goSaferEvaluationPackagesGosaferAccuracy{} on
the set of packages.
Its high precision allows it to be used productively.
Thus, \toolSafer{} is a valuable tool that can be added next to existing tools like \toolVet{} and \toolGosec{}, which
are not good at detecting \unsafe{} misuses with a large number of false negatives and false positives, respectively.
Using \toolSafer{}, \numberBugsFixed{} bugs in open-source Go libraries were found.
To fix them, \numberPRs{} pull requests have been submitted to the authors, and \checkNum{one} already existed.
So far, \numberPRsMerged{} pull requests have been merged, fixing \numberBugsMerged{} (\fractionBugsMerged{}) of the
bugs.

In summary, \unsafe{} code is commonly used in the most popular Go projects for a number of reasons such as efficiency
or the foreign function interface.
Using the novel static analysis tools contributed in this thesis, developers can embrace this fact and mitigate the
risks that come with it by effectively localizing and checking \unsafe{} usages in their own and third-party code.
Additionally, the novel evidence of how to actually exploit possible vulnerabilities related to \unsafe{} code helps
developers to understand the dangers and, ultimately, avoid them.
