\chapter*{Acknowledgments}

This thesis was written at Technische Universität Darmstadt under the supervision of Prof. Dr.-Ing. Mira Mezini,
Dr. rer. nat. Lars Baumgärtner, and Anna-Katharina Wickert, M.Sc.
Parts of this work have been submitted as a research paper titled "Uncovering the Hidden Dangers: Finding Unsafe Go Code
in the Wild", which is accepted at the 19th IEEE International Conference on Trust, Security and Privacy in
Computing and Communications (IEEE TrustCom 2020) \jl{add ArXiV citation}.
I explicitly acknowledge the contributions the co-authors Lars Baumgärtner, Anna-Katharina Wickert, and Mira Mezini
have made to that paper.
I made sure to properly cite all relevant prior and concurrent work as usual with the standards on scientific ethics
to clearly show which contributions are derived from the work of other people and which are mine.

I would like to thank Lars, Anna, and Mira for their excellent work supervising this thesis.
I could not imagine having had any better support during the past months.
Especially, I want to thank Anna for restoring my motivation in challenging times, and Lars for his honest and
demanding, but always very helpful feedback.

In addition, I want to thank my family for giving me both moral and financial support throughout my entire studies,
always believing in me and having my back, and my girlfriend for cheering me up even in the most stressful and hard
days.
In fact especially in those days.
Finally, a big thank you to my friends who have made the past years at university the great time that it has been, Timo
for letting me switch offices for a couple of days, and \todo{X}, \todo{Y}, and \todo{Z} for proof-reading this thesis.
